---
title: "Course project 1"
author: "ALE Junior"
date: '2022-06-15'
output: html_document
---

## Loading and preprocessing the data
 Load the data

```{r, echo=TRUE}
activityData <- read.csv(file="activity.csv", header=TRUE)
```

 No need to tranform the data

## What is mean total number of steps taken per day?
 Calculate the total steps taken per day
 
```{r, echo=TRUE}
totalSteps <- aggregate(steps ~ date, activityData, FUN=sum)
```

 Make a histogram of the total number of steps taken per day
 
```{r, echo=TRUE}
hist(totalSteps$steps,
     main = "Total Steps per Day",
     xlab = "Number of Steps")
```

 Calculate and report the mean and median of total steps taken per day
 
```{r, echo=TRUE}
meanSteps <- mean(totalSteps$steps, na.rm = TRUE)
medSteps <- median(totalSteps$steps, na.rm = TRUE)
```

## What is the average daily activity pattern?

 Make a time-series plot of the 5-minute interval and the average number of steps taken, averaged acoss all days
 
```{r, echo=TRUE}
library(ggplot2)
meanStepsByInt <- aggregate(steps ~ interval, activityData, mean)
ggplot(data = meanStepsByInt, aes(x = interval, y = steps)) +
  geom_line() +
  ggtitle("Average Daily Activity Pattern") +
  xlab("5-minute Interval") +
  ylab("Average Number of Steps") +
  theme(plot.title = element_text(hjust = 0.5))
```

 Which 5-minute interval across all days contain the maximum number of steps
```{r, echo=TRUE}
maxInt <- meanStepsByInt[which.max(meanStepsByInt$steps),]
```

## Imputing Missing Values

 Calculate and report the total number of missing values in the dataset
 
```{r, echo=TRUE}
missingVals <- is.na(activityData$steps)
table(missingVals)
```

 Devise a strategy for filling in all of the missing values

There are 2304 missing values. I will replace these missing values with the 5-day average of that respective interval.

 Create a new dataset that is equal to the original dataset but with the missing data filled in.
 
```{r, echo=TRUE}
new_activityData <- transform(activityData,
steps = ifelse(is.na(activityData$steps), meanStepsByInt$steps[match(activityData$interval, meanStepsByInt$interval)], activityData$steps))
```

 Make a histogram of the total number of steps taken each day and report the mean and median.
 
```{r, echo=TRUE}
impStepsByInt <- aggregate(steps ~ date, new_activityData, FUN=sum)
hist(impStepsByInt$steps,
     main = "Imputed Number of Steps Per Day",
     xlab = "Number of Steps")

impMeanSteps <- mean(impStepsByInt$steps, na.rm = TRUE)
impMedSteps <- median(impStepsByInt$steps, na.rm = TRUE)
diffMean = impMeanSteps - meanSteps
diffMed = impMedSteps - medSteps
diffTotal = sum(impStepsByInt$steps) - sum(totalSteps$steps)
```

There is a difference of 0 in the mean steps of the two dataset. There is a difference of -1.076381110^{4} in the median steps of the two dataset. There is a difference of 8.612950910^{4} in the total steps of the two dataset.

## Are there differences in activity patterns between weekdays and weekends?

 Create a new factor variable in the dataset with two levels - "weekend" and "weekday"
 
```{r, echo=TRUE}
DayType <- function(date) {
  day <- weekdays(date)
  if (day %in% c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday'))
      return ("weekday")
  else if (day %in% c('Saturday', 'Sunday'))
      return ("weekend")
  else
      stop ("Invalid Date Format.")
}
new_activityData$date <- as.Date(new_activityData$date)
new_activityData$day <- sapply(new_activityData$date, FUN = DayType)
```

 Make a panel plot containnig a time-series plot of the 5-minute interval and the average number of steps taken across all weekdays or weekends

```{r, echo=TRUE}
meanStepsByDay <- aggregate(steps ~ interval + day, new_activityData, mean)
ggplot(data = meanStepsByDay, aes(x = interval, y = steps)) + 
  geom_line() +
  facet_grid(day ~ .) +
  ggtitle("Average Daily Activity Pattern") +
  xlab("5-minute Interval") +
  ylab("Average Number of Steps") +
  theme(plot.title = element_text(hjust = 0.5))
```
